---
title: "R Notebook"
output: html_notebook
---

```{r}
# run all code chunks in this document
# do not display output of this codechunk
# DO display code
rm(list=ls())
options(scipen=999)
source("helper_plot_model.R")
source("helper_init_model.R")
source("helper_calc_mse.R")
source("helper_init_large_model.R")
library(asp20boost)
library(gamboostLSS)
library(microbenchmark)
```

### ToDo Sebastián
* Formula in front of every 'Configuration' paragraph
* Make Data Frame look like a nice table


## Small Model - Two-dimnesional Case

The goal in this first part is to demonstrate, that the algorithm finds correct estimates for sensible choices of *beta* and *gamma.* We use the following 4 different parameter configurations to explore location-scale regression model characteristics.

```{r}
# dont display code
# dont display output
# config 1 ----------------------
beta_1 <- c(0,1)
gamma_1 <- c(-3,2)

# config 2 ----------------------
beta_2 <- c(2,2)
gamma_2 <- c(2,2)

# config 3 ----------------------
beta_3 <- c(0,1)
gamma_3 <- c(-1,-0.3)

# config 4 ----------------------
beta_4 <- c(0,15)
gamma_4 <- c(1,-1)
```


```{r}
# display code
# dont display output
set.seed(1337)
mod_1 <- init_model(beta_1, gamma_1, 1000)$model
gradient_boost(mod_1)
mod_2 <- init_model(beta_2, gamma_2, 1000)$model
gradient_boost(mod_2)
mod_3 <- init_model(beta_3, gamma_3, 1000)$model
gradient_boost(mod_3)
mod_4 <- init_model(beta_4, gamma_4, 1000)$model
gradient_boost(mod_4)
```


```{r, cache=TRUE, fig.height = 8, fig.width = 11}
# display code
# display output
par(mfrow = c(2,2))
plot_model(beta_1, gamma_1, 1000, "Config 1")
plot_model(beta_2, gamma_2, 1000, "Config 2")
plot_model(beta_3, gamma_3, 1000, "Config 3")
plot_model(beta_4, gamma_4, 1000, "Config 4")
```
```{r}
# dont display code
# display output
result_df <- data.frame("Config" = c(1,1,2,2,3,3,4,4),
                        "beta_true" = c(beta_1, beta_2, beta_3, beta_4),
                        "beta est." = c(round(mod_1$beta, 3), round(mod_2$beta, 3), round(mod_3$beta, 3), round(mod_4$beta, 3)),
                        "gamma_true" = c(gamma_1, gamma_2, gamma_3, gamma_4),
                        "gamma est." = c(round(mod_1$gamma, 3), round(mod_2$gamma, 3), round(mod_3$gamma, 3), round(mod_4$gamma, 3))
                        )
result_df
```



### Evaluation Configuration 1

[insert model config by formula]

This parameter configuration is equivalent to the example case of the *asp20model* package. 
As already mentioned, heteroskedasticity is present and indicated by the funnel shape of the point cloud.
The solid black line represents the true response mean, while the dashed lines represent the true upwards and downwards standard deviations from this mean.
The red arrow indicates the standard deviation at the point x = 1,  that is the upper limit of covariate values, since covariates are design to be on the unit interval.
For cases with increasing scale, the red arrow hence indicates maximum standard deviation of the response.
The blue arrow depicts the slope of the location parameter vector *beta_1.*
The ratio of those two quantities is essential for estimability of the model. 
Here the blue arrow is dominant, hence the model may well be estimated. 
Estimates are displayed in the table above. 
We see that both *betahat* and *gammahat* estimates are close to the true parameter vectors *beta* and *gamma.*


#### Evaluation Configuration 2

Configuration 2 is a somewhat naive parameter configuration, where all parameters are set to value 2. 
Estimating this model does not yield valid results.
The gradient_boost() function runs into numerical problems.
To understand why this happens we need to understand the model characteristics resulting from parameter configuration 2-2-2-2.
Here, the red arrow is of large size, while the blue arrow is hardly visible. 
This is due to the presence of huge standard deviations in response *y*, concealing changes in mean of *y.* 
In fact, the standard deviation at x = 1 is exp(2+2) = 54.60.
In contrast the true response mean values range from 0 to 2.
The change in response location (the signal), indicated by the blue arrow, perishes due to the high amount of noise caused by the high gamma-parameters. 
Resulting is a signal-noise ratio, that is impossible to estimate. 
Values for parameters gamma need to be chosen sensibly for simulating data. 
This is no straight-forward task, since the effect of *gamma* on the response standard deviation is not linear, but exponential. 
To check this in the two-dimensional case, the arrows in the model plot are of help.


#### Evaluation Configuration 3

[insert parameter configuration as formula]

Configuration 3, in turn, yields reasonable estimates.
It contains a negative gamma slope, leading to decreasing response standard deviations.
This can be observed by the funnel shape pointing in the opposite direction as before.

It may be of interest to check if estimates lie closer to the true parameters than e.g. in configuration 1.
To do this we measure the gap between true values and estimates by the mean squared error.
This enables comparison of configurations 1 and 3.

```{r}
# display code
# display output
mse_1 <- calc_MSE(beta_1, gamma_1, mod_1$beta, mod_1$gamma)
mse_3 <- calc_MSE(beta_3, gamma_3, mod_3$beta, mod_3$gamma)
cbind("config 1" = mse_1,
      "config_3" = mse_3)
```

Configuration 1 leads to smaller MSE, hance our algorithm performs better in configuration 1.

#### Evaluation Configuration 4

[include configuration as formulas]

In configuration 4, we again included a negative gamma slope, as well as a very high beta slope.
The idea is to ensure a very good signal noise ratio, in order to ensure good model fit, indicated by low MSE.
Contrary to our expectations, the algorithm performs poorly for this parameter configuration.

```{r}
# display code
# display output
mse_4 <- calc_MSE(beta_4, gamma_4, mod_4$beta, mod_4$gamma)
cbind("config 1" = mse_1,
      "config_3" = mse_3,
      "config 4" = mse_4)
```

The bad model fit is due to the fact, that the location slope is highly underestimated.


### Conclusions

The observed two-dimensional cases let us conclude that not every arbitrary parameter configuration is estimable by the implemented gradient_boost() algorithm. A first insight is that signal-noise ratio needs to be sufficiently high in order to prevent that location effect gets concealed by inflated standard deviations. For the finding of configuration 4 we have several explanation approaches:
* There are further constraints on simulating data meaningfully which we were not yet able to work out.
* Our algorithm does not reliably yield robust results if negative gamma slopes are included.
* Our algortihm runs into problems if the signal is too strong compared to noise.
